\LARGE \bf Approximately Orchestrated Road Traffic Automata\\ a General Traffic Simulation Framework

### 2012-02-22 papers/itsc_2012/draft2.tex
worker threads dont do dumb work, deterministic cycled light policy with bad math
simulation section revised heavily

AORTA's implementation uses Scala, a language implemented on the Java Virtual
Machine. Scala provides the advantage of functional programming constructs while
still permitting imperative style. Parts of AORTA, including extensions and clients, can
just as easily be written in Java. The software has a documented API, is open-source, and
contains all of the code for the experiments described in Section 6 -- it is
designed for extensibility\footnote{http://code.google.com/p/road-rage}.

@@

The map also has a geometry interpretation, used for visualization and physical
distances. A road has a sequence of points giving the line segments of the
center yellow line. These points could also be interpreted as control points for
a curved road, but AORTA uses straight lines as an approximation. Lanes are
physically placed a lane's width away from and parallel to the center line. (In
the case of one-ways, there is no center line; instead, the lanes simply cluster
around the invisible center lane, all pointing the same way.) Turns are modeled
with a single straight line connecting the end of one edge to the start of
another (so right turns often have $0$ length).

@@

Map construction is split into several passes, each feeding the next.
\begin{enumerate}
  \item OSM encodes many paths besides driveable roads, so these are first
        filtered out. GPS locations common to multiple roads are marked, since
        they implicitly indicate intersections.
  \item OSM \emph{ways} (sequences of GPS locations) include many
        intersections, so the builder next splits ways into undirected
        segments of roads between exactly two intersections.
  \item Undirected roads are multiplied into directed lanes in each direction
        (unless the road is marked one-way). The number of lanes is guessed from
        OSM's ``road type'' tag or an explicit number of lanes, when given.
  \item Turns are next constructed between incoming and outgoing edges at each
        vertex. Left, right, straight, and U-turns are heuristically chosen
        based on the angles between edges. When several lanes all cross into
        fewer lanes, merging is arbitrarily forced on the rightmost lanes.
        Although inaccurate, these patterns are the best compromise. At the
        time of writing, even sources like Google Maps (TODO cite) lack this
        information.
  \item The strongly connected components of the digraph with the fewest
        members are removed. Tarjan's algorithm locates these in linear time
        (TODO cite). Disconnected lanes are spliced out of larger roads, leaving
        unappealing geometric gaps, but correctly ensuring connectivity for
        corner-case free pathfinding.
  \item The builder serializes the graph in a simple XML format.
\end{enumerate}

@@

AORTA uses microscopic simulation, modeling individual drivers as agents.
Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
position and velocity. The tick duration $dt$ is configurable and fixed
\commentpk{configurable \textit{and} fixed is confusing} for the simulation.
Agents know this value and use it to reason about collision avoidance and
satisfying other constraints, so setting this value to the average ``reaction
time'' of a human driver may be appropriate. Each tick, the simulation performs
the following steps:

\begin{enumerate}
  \item Introducing new agents into the map
  \item Updating the position and velocity of agents
  \item Checking for collisions
  \item Allowing each agent to react
\end{enumerate}

Agents have a few enforced constraints -- they cannot exceed their physical
acceleration capabilities, travel in reverse, or travel past the end of a
traversable without appearing on the next. Intersection policies and avoiding
collisions on the same traversable are the source of the other constraint in the
simulation, but it is the job of a behavior to satisfy those. There are three
main components of the simulation, each configurable by extending default
classes: agent behaviors, routing strategies, and intersection policies. Each
is described below.

\commentd{justify arbitrary constants, like deaccel for cars;
          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
          issues;
          limit: no lane-changing}

@@

\commentpk{I wonder if demand data generation in the ui should go here.}

To introduce new agents into the simulation in real-time, users create
\emph{generators} programatically or using the UI. A generator picks a random
starting position and random goal from two sets of roads chosen by the user. A
generator may cover the entire map, uniformly distributing traffic everywhere,
or it can mimic rush-hour scenarios. Every tick, a generator creates some number
of new agents.  If the agent's route policy needs work to be done (such as
pathfinding), the generator either performs it immediately (incurring a delay in
simulation if it is running) or sends it to a pool of worker threads to compute
in the background. When a generator finishes the work for an agent, it promotes
it to the ``ready'' state.

Once an agent's route is ready, the agent waits alongside its starting edge (as
if they are in a driveway or parking lot). At the beginning of each tick, they
enter with zero initial speed, if it is safe to do so. The list of agents
currently on the edge determines safety. Because behaviors must enforce speed
limit along an edge by the time an agent enters the road, then if there is no
agent within the worst-case distance (travel distance at the speed limit, plus
stopping distance from this speed) before the spawn point, then the new agent
may enter. To render unnecessary the prediction of which agents could enter the
edge in the near future, there is a buffer space at the beginning of the edge
where a new agent may never spawn. This disqualifies some short edges from being
spawn points, but this is not a big limitation in practice.

@@

There are two structures to detect collisions. All traversables (edges and the
straight line-approximation of turns) maintain a \emph{queue} of occupying
agents, ordering them so that the agent farthest along is at the head of the
queue. A naive method of collision detection would detect an agent passing an
agent on the same queue, or entering a new queue not at the end, immediately
after an agent moves. However, agents take their steps sequentially and in an
effectively random order. Ideally they could be reordered so that agents in the
front of a queue always move first. Since the first agent may cross into the
next queue, that introduces a dependency between traversables. A topological
sort of a directed graph containing circuits\commentp{do you mean cycles?} is not possible though, and since
we guarantee connectedness, there are necessarily circuits in maps.  Instead,
collisions are checked after every agent has stepped forward in time, treating
each time-step as atomic. Queues remember the original order of agents before a
step, and afterwards can verify that the order is still legal, with new agents
on the end and existing agents in the same order.\commentp{I don't
  follow the point of the last sentence.}

Likewise, intersections check for collisions after everybody has moved by
examining the turn each agent is performing and verifying that no other agent is
simultaneously performing a conflicting turn.

\commentpk{The following paragraph should not be here. It should be in intersection behaviors} AORTA has a low-granularity model of turn conflict. If two vehicles at any point
along two different turns could ever come into contact, then the turns are
always in conflict, no matter where the agent specifically is along that turn.
The alternative of tiling intersections (like \cite{JAIR08-dresner} \commentpk{I made this same mistake a month ago, dont use citations as a noun} does) would
provide more granularity, but it adds computational complexity and would be
impractical for many of the complex intersection geometries inferred from OSM.

@@

\commentp{What happens when a collision is detected? }

@@

Each agent has a ``behavior'' governing it. When the agent travels past the end
of an edge during its step, the behavior picks the turn to pursue. Each tick,
the behavior can make the agent react by performing one of two actions:
disappearing from the map (only when the agent is at rest and is done with its
route) or setting an acceleration for the next step. In the future, another
possible action will be initiating a lane-change or
merge.\commentp{Say ``Were lane changing or merging modeled, that
  would be another possible action.''}

Currently, the primary behavior used by all agents is the ``route-following
behavior.'' It is a generalized, baseline behavior that guarantees not to
collide with another agent or enter an intersection at the wrong time. Since it
picks the fastest conservative (safe) choice of acceleration at each step, it
could be easily extended to mimic human drivers by traveling at some random
speed slower than this or to optimize fuel efficiency by tuning acceleration.

@@

Since agents know the fixed $dt$ duration, they may reason about what could
happen during the next tick to avoid collision and illegal intersection entry.
The conservative analysis is to determine worst case -- how far could an agent
travel by the end of the next tick? If they accelerate as fast as possible for
$dt$, this gives them the longest possible traveling distance and highest
possible speed. Adding the stopping distance for this speed to the traveling
distance gives the worst-case distance.

\commentd{how much formula? pretty simple so far, but can certainly include them.
 should this part be worded like a proof? (because it can be)}
\commentp{Use a formula if you can use it to mkae the description more
  concise (and precise).}

The worst-case distance may exceed the distance remaining on the agent's current
traversable. That means they could start or finish a turn during the next tick.
Because edges are often extremely short due to mis-marked ways in OSM (less than
$???$ meters), they may travel through several traversables during the next
move. Thus, any conservative behavior must look ahead until the worst-case
distance is exhausted, asking the route strategy which turn an agent will choose
in the future. At each traversable considered, there are 3 constraints to
satisfy: obeying the intersection, avoiding collision with another agent, and
obeying speed limits. Each constraint contributes to the max acceleration that
will obey the limit, so ultimately the result of lookahead will be the minimum
of these candidates.

Obeying a road's speed limit is a simple constraint. Like many parameters of the
map, speed limits are inferred from the road type claimed by the OSM
source.

\commentp{This section seems belabored.  Can't you just say that the
  agent sets its acceleration to be the maximum possible that satifies
  3 constraints: staying at or below the speed limit, avoiding
  collisions, and not entering an intersection illegally.}

\subsubsection{Obeying intersections}

A behavior's interaction with an intersection policy amounts to polling it with
the turn the agent wants to perform, the distance away the agent currently is,
and how long the agent has been waiting (so policies can enforce a required
delay before entering the intersection). The intersection policy simply orders
the agent to ``stop'' or ``continue.'' The lookahead considers no intersections
beyond the first that issues ``stop'', since the behavior cannot legally proceed
past that one anyway. The acceleration to avoid entering an intersection
attempts to end a configurable threshold back from the end of a traversable to
account for crosswalks and as an epsilon value, since achieveing a distance
along an edge equal to its length means spilling over into the next traversable,
which the intersection has banned. Sometimes stopping behind this threshold is
not possible -- namely, when the edge length is shorter -- and in that case,
anywhere before the end is acceptable.

\commentd{include some form of piyush's proof about accel to end?} \commentpk{I would say it is unnecessary}

\subsubsection{Optimal tailgating}

\commentd{diagram with overlapping distances}

When the lookahead analysis is considering some future traversable, it only has
to consider the agent closest to the start of it in the present. An edge may
only be entered from a turn, so the intersection policy has the burden of
ensuring no agent performs a turn that leads to the same edge simultaneously.
Following an agent safely means maintaining a configurable following distance
behind the agent \commentpk{not sure what you mean by configurable here}. Worst-case analysis is again used: the most dangerous move the
agent in front could perform is decelerating as fast as possible
during the next tick,
and the worst-case traveling distance for the agent following is already known
from lookahead (speeding up as much as possible).

@@

\commentd{diagram showing deadlock, or lookahead in gen eral}
\commentd{i really hope to fix this issue soon, but it's not an immediately easy
fix}
\commentp{Cite (and ideally use ideas from) paper commented out in latex file
    % @inproceedings{AAAI11-au,
    % Author = {Tsz-Chiu Au and Neda Shahidi and Peter Stone},
    % Booktitle= {Proceedings of the Twenty-Fifth Conference on Artificial Intelligence},
    % Title = {Enforcing Liveness in Autonomous Traffic Management},
    % Year = {2011},
    % Month= {August},
}

Picking accelerations that optimize some objective functions is a separate task
from picking a path to travel. Thus, the behavior described above can consult
any route strategy. Currently two simple implementations are available. One
statically plans a route to some goal using standard A* search; the other
performs a drunken walk, lazily planning random choices at each intersection as
the lookahead engine needs to know. Future experiments in dynamic replanning or
hierarchial planning can be performed without modifying any code other than a
creating new route implementation. Route policies could also interact with a
singleton overseer to have some sort of global insight.

% TODO and mike's variant that wanders in the right direction

\subsection{Intersection Policies}

Agents and behaviors (on behalf of an agent) poll an intersection as they
approach it, asking if they should stop or not. A policy is free to prevent
agents from performing conflicting turns concurrently by any means. Two main
policies exist already, but implementing the interface for extensions is simple.

The stop sign policy works by maintaining a current owner and a queue of waiting
agents. An agent can enter the queue only when it is within some threshold of
the start of its turn (a constraint satisfied by the default lookahead
behavior), even if that is several (rather short) edges and turns away. After
the agent at the head of the queue has waited for a configurable pause, it obtains the lock
and has exclusive access to the intersection, releasing it only when it
completes its turn.

A form of deadlock is possible in a sequence of short edges and turns. Agent
$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
To deal with this circumstance, an agent's behavior can yield any held locks (a
notion generalized to any intersection policy) when its speed is $0$ and its
conservative acceleration is to remain at rest. The stop sign policy only
removes an agent's lock if it has not started its turn (because otherwise, the
agent is stalling in the intersection). The next time that agent polls, the
policy restores the agent to the front of the queue.

\commentd{hopefully we'll have a neater solution than the above soon}

\commentd{describe traffic signal policy, once it's working}

Intersection policies are one of the most intriguing aspects of the simulation to
tweak, especially in light of autonomous vehicles. Future policies could include stop lights that adaptively learn better
timings and an AIM-based reservation system \cite{JAIR08-dresner}.

>>>>

AORTA's implementation uses Scala, a language implemented on the Java Virtual
Machine. Scala provides the advantage of functional programming constructs while
still permitting imperative style. Parts of AORTA, including extensions and clients, can
just as easily be written in Java. The software has a documented API, is
open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
the code for the experiments described in Section 6 -- it is designed for
extensibility.

@@

The map also has a geometric interpretation, used for visualization and physical
distances. A road has a sequence of points giving the line segments of the
center yellow line. These points could also be interpreted as control points for
a curved road, but AORTA uses straight lines as an approximation. Lanes are
physically placed a lane's width away from and parallel to the center line. (In
the case of one-ways, there is no center line; instead, the lanes simply cluster
around the invisible center lane, all pointing the same way.) Turns are modeled
with a single straight line connecting the end of one edge to the start of
another (so right turns often have $0$ length).

@@

Map construction is split into several passes, each feeding the next.
\begin{enumerate}
  \item OSM encodes many paths besides driveable roads, so these are first
        filtered out. GPS locations common to multiple roads are marked, since
        they implicitly indicate intersections.
  \item OSM \emph{ways} (sequences of GPS locations) include many
        intersections, so the builder next splits ways into undirected
        segments of roads between exactly two intersections.
  \item Undirected roads are multiplied into directed lanes in each direction
        (unless the road is marked one-way). The number of lanes is guessed from
        OSM's ``road type'' tag or an explicit number of lanes, when given.
  \item Turns are next constructed between incoming and outgoing edges at each
        vertex. Left, right, straight, and U-turns are heuristically chosen
        based on the angles between edges. When several lanes all cross into
        fewer lanes, merging is arbitrarily forced on the rightmost lanes.
        Although inaccurate, these patterns are the best compromise. At the
        time of writing, even sources like Google Maps (TODO cite) lack this
        information.
  \item Because turns are heuristically placed, the graph is disconnected. The
        strongly connected components of the digraph with the fewest members are
        removed. Tarjan's algorithm locates these in linear time (TODO cite).
        Disconnected lanes are spliced out of larger roads, leaving unappealing
        geometric gaps, but correctly ensuring connectivity for corner-case free
        pathfinding.
  \item The builder serializes the graph in a simple XML format.
\end{enumerate}

@@

AORTA uses microscopic simulation, modeling individual drivers as agents.
Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
position and velocity. The tick duration $dt$ is fixed, enabling agents to
reason about collision avoidance and satisfying other constraints. Setting this
value to the average ``reaction time'' of a human driver may be appropriate for
some experiments. Each tick, the simulation performs the following steps:

\begin{enumerate}
  \item Introducing new agents into the map
  \item Updating the position and velocity of agents based on their choice the
        previous tick
  \item Checking for collisions
  \item Allowing each agent to react
\end{enumerate}

Agents have a few enforced constraints -- they cannot exceed their physical
acceleration capabilities, travel in reverse, or travel past the end of a
traversable without appearing on the next. Intersection policies and avoiding
collisions on the same traversable provide the other constraints, but it is the
job of an agent's behavior to satisfy those. There are three main components of
the simulation, each configurable by extending default implementations of an
interface: agent behaviors, routing strategies, and intersection policies.

%\commentd{justify arbitrary constants, like deaccel for cars;
%          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
%          issues;
%          limit: no lane-changing}

@@

To introduce new agents into the simulation in real-time, users create
\emph{generators} programatically or using the UI. A generator picks a random
starting position and random goal from two sets of roads chosen by the user,
selectable by drawing a polygon with the mouse in the UI. A generator may cover
the entire map, uniformly distributing traffic everywhere, or it can mimic
rush-hour scenarios.

Every tick, a generator creates some number of new agents. If the agent's route
policy needs work to be done (such as pathfinding), the generator either
performs it immediately (incurring a delay in simulation if it is running) or
sends it to a pool of worker threads to compute in the background.

Once an agent's route is ready, the agent waits alongside its starting edge (as
if it is in a driveway or parking lot). When the edge has no agents that could
potentially crash with the new vehicle, the new agent enters the system with
zero initial speed.  Because behaviors must enforce speed limit along an edge by
the time an agent enters the road, then if there is no agent within the
worst-case distance (traveling distance at the speed limit, plus stopping
distance from this speed) before the spawn point, then the new agent may enter.
To render unnecessary the prediction of which agents could enter the edge in the
near future, there is a buffer space at the beginning of the edge where a new
agent may never spawn. This disqualifies some short edges from being spawn
points, but this is not a big limitation in practice.

@@

All traversables (edges and the straight line-approximation of turns) maintain
an ordered \emph{queue} of occupying agents. Agents stepping forward in time
happens atomically, so collision checking occurs after all agents have moved for
some tick. This is done because otherwise false positives could occur when an
agent steps before the agent in front of it, and ordering agents for updates is
difficult near the crossings between edges. \commentd{Do I need to explain why
it is difficult, or should I just stop after ``in front of it''?} A collision is
defined as the reversal of order of agents before and after a step.

Likewise, intersections check for collisions by examining the turn each agent is
performing and verifying that no other agent is simultaneously performing a
conflicting turn.  AORTA has a low-granularity model of turn conflict. If two
vehicles at any point along two different turns could ever come into contact,
then the turns are always in conflict, no matter where the agent specifically is
along that turn.  The alternative of tiling intersections \cite{JAIR08-dresner}
would provide more granularity, but it adds computational complexity and would
be impractical for many of the complex intersection geometries inferred from
OSM.

When a collision occurs, the simulator throws warnings or halts, since this
indicates a bug in some policy. A future alternative could be to mark the region
as impassable and re-route other traffic away from the accident.

@@

@@

Each agent has a ``behavior'' governing it. When the agent travels past the end
of an edge during its step, the behavior picks the turn to pursue. Each tick,
the behavior can make the agent react by performing one of two actions:
disappearing from the map (only when the agent is at rest and is done with its
route) or setting an acceleration for the next step. Were lane changing or
merging modeled, that would be another possible action.

Currently, the primary behavior used by all agents is the ``route-following
behavior.'' It is a generalized, baseline behavior that guarantees not to
collide with another agent or enter an intersection at the wrong time. Since it
picks the highest safe choice of acceleration at each step, it could be easily
extended to mimic human drivers by traveling at some random speed slower than
this or to optimize fuel efficiency by tuning acceleration.

@@

\commentd{one diagram probably covers lots of this}

The default behavior maximizes accelerations while obeying speed limits,
avoiding collisions, and obeying intersections. By consdering the highest speed
achievable at the end of the next tick and the subsequent stopping distance at
this speed, the behavior can scan sufficiently far ahead along several
traversables to locate agents to avoid and intersections to query. The analysis
can short-circuit at the first nearby agent and intersection requiring a stop,
since it would be illegal to pass either.

\subsubsection{Interaction with intersections}

% TODO regroup? does this belong as a subsection, not subsub?

A behavior's interaction with an intersection policy amounts to polling it with
the turn the agent wants to perform, the distance away the agent currently is,
and how long the agent has been waiting (so policies can enforce a required
delay before entering the intersection). The intersection policy simply orders
the agent to ``stop'' or ``continue.'' The acceleration to avoid entering an intersection
attempts to end a configurable threshold back from the end of a traversable to
account for crosswalks.

Intersection policies are one of the most intriguing aspects of the simulation
to tweak, especially in light of autonomous vehicles. Current policies include
traditional stop signs and traffic signals. Future policies could include stop
lights that adaptively learn better timings and an AIM-based reservation system
\cite{JAIR08-dresner}.

\commentd{Actually, at least the stop sign policy is really mundane. Include its
  description?}

%The stop sign policy works by maintaining a current owner and a queue of waiting
%agents. An agent can enter the queue only when it is within some threshold of
%the start of its turn. After the agent at the head of the queue has waited for a
%configurable pause, it obtains the lock and has exclusive access to the
%intersection, releasing it only when it completes its turn.

\commentd{describe traffic signal policy, once it's working fully}

Picking accelerations that optimize some objective functions is a separate task
from picking a path to travel. Behaviors consult a route strategy to pick turns.
Currently two simple implementations are available. One statically plans a route
to some goal using standard A* search; the other performs a drunken walk, lazily
planning random choices at each intersection as the lookahead engine needs to
know. Future experiments in dynamic replanning or hierarchial planning can be
performed without modifying any code other than a creating new route
implementation. Route policies could also interact with a singleton overseer to
have some sort of global insight.

\subsection{Deadlock}

@@

%\commentp{Cite (and ideally use ideas from) paper commented out in latex file
    % @inproceedings{AAAI11-au,
    % Author = {Tsz-Chiu Au and Neda Shahidi and Peter Stone},
    % Booktitle= {Proceedings of the Twenty-Fifth Conference on Artificial Intelligence},
    % Title = {Enforcing Liveness in Autonomous Traffic Management},
    % Year = {2011},
    % Month= {August},
%}

% TODO solve the above case, then figure out what to describe.

%A form of deadlock is possible in a sequence of short edges and turns. Agent
%$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
%proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
%To deal with this circumstance, an agent's behavior can yield any held locks (a
%notion generalized to any intersection policy) when its speed is $0$ and its
%conservative acceleration is to remain at rest. The stop sign policy only
%removes an agent's lock if it has not started its turn (because otherwise, the
%agent is stalling in the intersection). The next time that agent polls, the
%policy restores the agent to the front of the queue.

%\commentd{hopefully we'll have a neater solution than the above soon}
